Design and Implementation of Probabilistic Programming Language Anglican

### 2016-07-04 paper/paper.tex
layout

\documentclass[a4paper]{article}

@@

\title{Notes on Design and Implementation of Anglican}
\author{David Tolpin\\dtolpin@robots.ox.ac.uk}

@@

Opaque, or value, expressions, have a different shape in the
original and the CPS form. However, their CPS form follows the
pattern \texttt{(continuation transformed-expression)}, and thus
the transformation does not depend on the continuation, and
can be accomplished without passing the continuation as a
transformation argument. Primitive (non-CPS) procedures used in
Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
forms are opaque and transformed by
\texttt{primitive-procedure-cps}, \texttt{fn-cps}, and
\texttt{mem-cps}, correspondingly.

@@

To realize potential danger of code size explosion, consider
CPS transformation of code
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
(if (adult? person)
  (if (male? person)
     (choose-beer)
     (choose-wine))
  (choose-juice))
\end{lstlisting}
with continuation 
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
(fn [choice _]
  (case (kind choice)
    :beer (beer-jar choice)
    :wine (wine-glass choice)
    :juice (juice-bottle choice)))
\end{lstlisting}
The code of the continuation, represented by an \texttt{fn}
form, will be repeated three times. In general, CPS code can
grow extremely large if symbolic continuations are inserted
repeatedly.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime, however this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is a Clojure function. Global dynamically
bound variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespace*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure function inside the let block in the
following example:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

@@

The sole purpose of the algorithm identifier is method dispatch
--- conventionally, the identifier is a keyword related to the
algorithm name (\texttt{:lmh} for Lightweight
Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
etc.). The second parameter is a query as defined by
\texttt{query} or \texttt{defquery} forms. If needed, the query
can be defined anonymously, right in the argument list of a call
to \texttt{infer}:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
(let [x 1]
  (infer :pgibbs (query (predict x)) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, however the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. The value is
destructured in case of a structured binding, for example:
\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
(defquery my-query [mean sd]
  (predict (sample (normal mean sd))))

@@

The simplest inference algorithm is \textit{importance
sampling}:
\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily, showstringspaces=false]
(derive ::algorithm :anglican.inference/algorithm)

(defmethod infer :importance [_ prog value & {}]
  (letfn [(sample-seq []
            (lazy-seq
              (cons
                (:state (exec ::algorithm prog value
                              initial-state))
                (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
\texttt{anglican.importance/infer} just calls
\texttt{anglican.inference/exec} and relies on default
implementations of checkpoint handlers. A different inference
algorithm would provide its own implementations of
\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
both, as well as invoke \texttt{exec} from an elaborated
conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
(\texttt{anglican.smc}) are examples of inference algorithms
where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
handler is overridden. In addition, SMC runs multiple particles
(program instances) simultaneously, while LMH runs re-runs
programs from an intermediate continuation rather than from the
beginning.

@@

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object
implements the \texttt{anglican.runtime/distribution} protocol,
with two methods: \texttt{sample} and \texttt{observe}. The
names of the methods coincide with Anglican special forms
\texttt{sample} and \texttt{observe}, but they are by no means
the same. The \texttt{sample} \textit{method} returns a random sample,
the \texttt{observe} \textit{method} returns the log probability
of the value. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample} or
the \texttt{observe} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or re-used.

@@

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures', are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/random-process} protocol. The protocol
has twho methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican.

\bibliographystyle{alpha}
\bibliography{refs}

>>>>

\documentclass[preprint]{sigplanconf}

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\title{Design and Implementation of Probabilistic Programming Language Anglican}
\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}

@@

Opaque, or value, expressions, have a different shape in the
original and the CPS form. However, their CPS form follows the
pattern \texttt{(continuation transformed-expression)}, and thus
the transformation does not depend on the continuation, and
can be accomplished without passing the continuation as a
transformation argument. Primitive (non-CPS) procedures used in
Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
forms are opaque and transformed by
\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps}, \texttt{fn-cps}, and
\texttt{mem-cps}, correspondingly.

@@

To realize potential danger of code size explosion, consider
CPS transformation of code
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(if (adult? person)
  (if (male? person)
     (choose-beer)
     (choose-wine))
  (choose-juice))
\end{lstlisting}
with continuation 
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(fn [choice _]
  (case (kind choice)
    :beer (beer-jar choice)
    :wine (wine-glass choice)
    :juice (juice-bottle choice)))
\end{lstlisting}
The code of the continuation, represented by an \texttt{fn}
form, will be repeated three times. In general, CPS code can
grow extremely large if symbolic continuations are inserted
repeatedly.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime, however this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is a Clojure function. Global dynamically
bound variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespace*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure function inside the let block in the
following example:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

@@

The sole purpose of the algorithm identifier is method dispatch
--- conventionally, the identifier is a keyword related to the
algorithm name (\texttt{:lmh} for Lightweight
Metropolis-Hasting, \texttt{:pcascade} for Particle Cascade
etc.). The second parameter is a query as defined by
\texttt{query} or \texttt{defquery} forms. If needed, the query
can be defined anonymously, right in the argument list of a call
to \texttt{infer}:
\begin{lstlisting}[frame=single,language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(let [x 1]
  (infer :pgibbs (query (predict x)) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, however the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. The value is
destructured in case of a structured binding, for example:
\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(defquery my-query [mean sd]
  (predict (sample (normal mean sd))))

@@

The simplest inference algorithm is \textit{importance
sampling}:
\begin{lstlisting}[frame=single, language=Anglican, basicstyle=\ttfamily\small, showstringspaces=false]
(derive ::algorithm
        :anglican.inference/algorithm)
(defmethod infer :importance [_ prog value & {}]
  (letfn
    [(sample-seq []
       (lazy-seq
         (cons
           (:state (exec ::algorithm prog value
                         initial-state))
           (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
\texttt{anglican.importance/{\linebreak[0]}infer} just calls
\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
implementations of checkpoint handlers. A different inference
algorithm would provide its own implementations of
\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
both, as well as invoke \texttt{exec} from an elaborated
conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
(\texttt{anglican.smc}) are examples of inference algorithms
where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
handler is overridden. In addition, SMC runs multiple particles
(program instances) simultaneously, while LMH runs re-runs
programs from an intermediate continuation rather than from the
beginning.

@@

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object
implements the \texttt{anglican.runtime/{\linebreak[0]}distribution} protocol,
with two methods: \texttt{sample} and \texttt{observe}. The
names of the methods coincide with Anglican special forms
\texttt{sample} and \texttt{observe}, but they are by no means
the same. The \texttt{sample} \textit{method} returns a random sample,
the \texttt{observe} \textit{method} returns the log probability
of the value. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample} or
the \texttt{observe} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or re-used.

@@

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures', are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
has twho methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican.

\acks

\bibliographystyle{abbrvnat}
\bibliography{refs}
